---
title: "Model"
output: html_document
---

#To do:

check condictions
and make initial model

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("betareg")
```


get packages. 
```{r}
library(tidyverse)
library(caTools)
library(gridExtra)
library(lmtest)
library(sandwich)
library(car)
library(caret)
library(nlme)
library(betareg )
set.seed(123)
```


```{r}
plot(lm.1)
```


need to get the high correlated values for each eda then we can do from there
```{r}
edu_corval <- c("School_spending", "Test_scores", "Colleges") #for edu stuff
social_cor <- c("Black", "Seg_racial", "Commute", "Gini", "Middle_class", 
               "Progressivity", "Manufacturing", "Migration_in", "Religious", 
               "Violent_crime", "Single mothers", "Divorced") #social stuff

high_cor <- mobility[, names(mobility) %in% c("Mobility",edu_corval, social_cor)]

```

```{r}
yikes <- lm(Mobility ~ ., data = high_cor)

plot(yikes)
```
get vif score for colinearity.
```{r}
vif(yikes)
```
nothing moajor above but may need to reasses if not good enough 
Do something call robust standard errors test that I found out to deal with that shit
```{r}
coeftest(yikes, vcov = vcovHC(yikes, type = "HC3"))
```

refitting after all this shit.
```{r}
yikes_refit <- lm(Mobility ~ Black + Seg_racial + Middle_class + Progressivity + Manufacturing + Migration_in + Religious + Divorced, data = high_cor)
coeftest(yikes_refit, vcov = vcovHC(yikes_refit, type = "HC3"))
```

```{r}
summary(yikes_refit)
```

```{r}
plot(yikes_refit )
```


```{r}
high_cor_clean <- high_cor[-c(320, 325, 326), ]
yikes_clean <- lm(Mobility ~ Black + Seg_racial + Middle_class + Progressivity + Manufacturing + Migration_in + Religious + Divorced, data = high_cor_clean)
summary(yikes_clean)
plot(yikes_clean)
```


Since what we are doing to trying to perdict a probability we need to do something the regresses off of a probability because DUH!! Kinda makes since right? Wish we learned more models instead of getting hit with some meaningless thoery that he said he would hide but does it anyway because he likes it. LIKE THE THOERY IS NOT EASY FOR UNDERGRADS..... But what do I know.... I just work here. never looked to see if it followed a beta distribution...


Does this look like a beta distribution
```{r}

hist(high_cor$Mobility, breaks = 30, probability = TRUE, main = "Histogram of Mobility", col = "lightblue")

```
Above not conclusive numeric test is needed
```{r}
library(MASS)

# Fit beta distribution parameters
mobility_fit <- fitdistr(high_cor$Mobility, "beta", start = list(shape1 = 2, shape2 = 5))

# Generate theoretical beta values
x_vals <- seq(min(high_cor$Mobility), max(high_cor$Mobility), length.out = 100)
beta_vals <- dbeta(x_vals, mobility_fit$estimate[1], mobility_fit$estimate[2])

# Plot the histogram and beta density curve
hist(high_cor$Mobility, breaks = 30, probability = TRUE, col = "lightblue", main = "Fit of Beta Distribution")
lines(x_vals, beta_vals, col = "red", lwd = 2)

ks.test(high_cor$Mobility, "pbeta", mobility_fit$estimate[1], mobility_fit$estimate[2])

```
p value > 0.5 so it is a beta distribution. shocking, now we can use beta regression.


```{r}
yikes_beta <- betareg(Mobility ~ Black + Seg_racial + Middle_class + Progressivity + Manufacturing + Migration_in + Religious + Divorced, 
                      data = high_cor)
summary(yikes_beta)
plot(yikes_beta)
```

comparing models... Seeing if black varies on middle class.
```{r}
yikes_interact <- betareg(Mobility ~ Black * Middle_class + Seg_racial + Progressivity + Manufacturing + Migration_in + Religious + Divorced, data = high_cor)
summary(yikes_interact)
plot(yikes_interact)
```







