---
title: "Project 1"
author: "Ruth Walters"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(ggplot2)
library(dplyr)
library(corrplot)
library(tidyr)
library(cowplot)
library(ggpubr)
library(GGally)
library(grid)
library(gridExtra)
library(MASS)
library(car)
library(boot)


theme_set(theme_bw())
theme_update(
  plot.title = element_text(size = 12, face = "italic"), 
  axis.title.x = element_text(hjust = 0),
  axis.title.y = element_text(angle=90, hjust = 0, margin = margin(r = 5))
)
```

```{r plotting functions}
plot_four <- function(a, b, c, d, title) {
  plot_row <- plot_grid(a,b,c,d, align = "hv")
  
  title <- ggdraw() + 
    draw_label(title,
               fontface = 'bold',
               x = 0,
               hjust = 0) +
    theme(plot.margin = margin(0, 0, 0, 7))
  
  plot_grid(title, 
            plot_row,
            ncol = 1,
            rel_heights = c(0.1, 1))
}

plot_two <- function(a, b, title) {
  plot_row <- plot_grid(a,b, align = "hv")
  
  title <- ggdraw() + 
    draw_label(title,
               fontface = 'bold',
               x = 0,
               hjust = 0) +
    theme(plot.margin = margin(0, 0, 0, 7))
  
  plot_grid(title, 
            plot_row,
            ncol = 1,
            rel_heights = c(0.1, 1))
}
```

# Introduction

Economic mobility, or the ability of an individual to raise their economic status throughout their lifetime, is a marker of a healthy society. As economic mobility declines and income inequality rises throughout the United States, it is of increasing interest to determine which factors contribute to immobility. In this paper, we will investigate the correlation between economic, educational, and policy factors that contribute to economic mobility. We hypothesize that economic factors such as income inequality, will be most predictive of economic mobility.

# Exploratory data analysis

```{r}
# Read in the mobility dataframe
mobility <- read.csv("mobility-all.csv", 
                     header = TRUE, 
                     stringsAsFactors = TRUE)
```

```{r view NAs, echo=TRUE}
# View NAs
nas <- colSums(is.na(mobility))
print(nas[nas > 0])
```


```{r clean data}
# Remove rows where a value for mobility is missing
mobilityTotalRows <- nrow(mobility)

mobility <- mobility %>%
  drop_na(Mobility) 

# Segment education-related variables
edu_mobility <- mobility[,(names(mobility) %in% c("Mobility", 
                                                   "School_spending", 
                                                   "Student_teacher_ratio", 
                                                   "Test_scores", 
                                                   "HS_dropout", 
                                                   "Colleges", 
                                                   "Tuition", 
                                                   "Graduation"))]

# Drop NAs
edu_mobility <- drop_na(edu_mobility)



# Remove qualitative variables
quals <- c("ID","Name", "State", "Latitude", "Longitude")
mobility <- mobility[,!(names(mobility) %in% quals)]

# Remove columns where more than 100 values are missing
bad_cols <- c("Colleges","Tuition", "Graduation", "HS_dropout") # +100 NULL
mobility <- mobility[,!(names(mobility) %in% bad_cols)]

# Drop remaining NA variables
mobility <- drop_na(mobility)

```

This dataset contains several rows for which one or more than one value is `NA`. Three steps were taken to eliminate `NA`s from the dataset while preserving its integrity.

1. *Drop the 12 rows that do not contain a value for `Mobility`* | These rows are useless for linear analysis because they do not contain the variable we are attempting to predict.

2. *Drop the features that contain a high incidence of `NA`s* | Any features that contained more than 100 NAs were designated as too poor in quality to be useful for the linear model. While some of these features were used for exploratory data analysis, they were removed from the dataset prior to modeling. 

3. *Drop all remaining `NA` values* | After removing the most `NA` values, a small amount of rows with `NA`s remained. These rows were dropped. 


Simply dropping all rows with `NA`s would have resulted in a reduction of `r (1 - (nrow(edu_mobility) / mobilityTotalRows)) * 100`% of the data whereas our three-step procedure only resulted in a reduction of `r (1 - (nrow(mobility) / mobilityTotalRows)) * 100`%.


Additionally, qualitative (non-numeric and non-quantitative) variables such as those representing latitude and longitude, state/region names, and the ID tag, were removed. 

## Education analysis

```{r educational investment}
# Plot student teacher ratio distribution
p1 <- ggplot(edu_mobility, aes(x= Student_teacher_ratio)) + 
  geom_histogram(bins = 30, fill = 'dodgerblue', color = "black") + 
  ggtitle("Student Teacher Ratio Distribution") +
  xlab("Student-teacher ratio") +
  ylab("Frequency") 

p2 <- ggplot(edu_mobility, aes(x = Student_teacher_ratio, y= Mobility)) + 
  geom_point(color = "dodgerblue", alpha = .3)  + 
  stat_smooth(method = "lm", formula = y ~ x, geom = "line", color = "black", se = FALSE) +
  stat_cor(label.x=18, label.y=.3) +
  ggtitle("Student Teacher Ratio vs Mobility") +
  xlab("Student-teacher ratio") +
  ylab("Mobility")

# Plot school spending distributions
p3 <- ggplot(edu_mobility, aes(x= School_spending)) + 
  geom_histogram(bins = 30, fill = 'tomato', color = "black") + 
  ggtitle("School Spending Distribution") +
  xlab("School spending ($)") +
  ylab("Frequency") + 
  scale_x_continuous(labels = scales::dollar_format())

p4 <- ggplot(edu_mobility, aes(x = School_spending, y= Mobility)) + 
  geom_point(color = "tomato", alpha = .3)  + 
  stat_smooth(method = "lm", formula = y ~ x, geom = "line", color = "black", se = FALSE) +
  stat_cor(label.x=7, label.y=.3) +
  ggtitle("School Spending vs Mobility") +
  xlab("School spending ($)") + 
  ylab("Mobility") + 
  scale_x_continuous(labels = scales::dollar_format())

#(p1 + p3) / (p2 + p4)
plot_four(p1,p3,p2,p4, title = "Educational investment as a predictor of economic mobility")
```



```{r educational outcomes}
# Plot dropout rates distribution
p1 <- ggplot(edu_mobility, aes(x= HS_dropout)) + 
  geom_histogram(bins = 30, fill = 'dodgerblue', color = "black") + 
  ggtitle("HS Dropout Distribution") +
  xlab("High school dropout rate") +
  ylab("Frequency")

p2 <- ggplot(edu_mobility, aes(x = HS_dropout, y= Mobility)) + 
  geom_point(color = "dodgerblue", alpha = .3)  + 
  stat_smooth(method = "lm", formula = y ~ x, geom = "line", color = "black", se = FALSE) +
  stat_cor(label.x=.025, label.y=.3) +
  ggtitle("HS Dropout vs Mobility") + 
  xlab("High school dropout rate") +
  ylab("Mobility")

# Plot test score distribution
p3 <- ggplot(edu_mobility, aes(x= Test_scores)) + 
  geom_histogram(bins = 30, fill = 'tomato', color = "black") + 
  ggtitle("Test Scores Distribution")+
  xlab("Test scores") +
  ylab("Frequency") 

p4 <- ggplot(edu_mobility, aes(x = Test_scores, y= Mobility)) + 
  geom_point(color = "tomato", alpha = .3)  + 
  stat_smooth(method = "lm", formula = y ~ x, geom = "line", color = "black", se = FALSE) +
  stat_cor(label.x=-20, label.y=.3) +
  ggtitle("Test Scores vs Mobility")+
  xlab("Test scores") +
  ylab("Mobility")

plot_four(p1,p3,p2,p4, title = "Educational outcomes as a predictor of economic mobility")
```



```{r tuition}
# Plot tuition distribution
p1 <- ggplot(edu_mobility, aes(x= Tuition)) + 
  geom_histogram(bins = 30, fill = 'dodgerblue', color = 'black') + 
  ggtitle("Tuition distribution") +
  xlab("Tuition costs ($)") + 
  ylab("Frequency") +
  scale_x_continuous(labels = scales::dollar_format())

p2 <- ggplot(edu_mobility, aes(x = Tuition, y= Mobility)) + 
  geom_point(color = "dodgerblue", alpha = .3) + 
  stat_smooth(method = "lm", formula = y ~ x, geom = "line", color = "black", se = FALSE) +
  stat_cor(label.x=11000, label.y=.3) +
  ggtitle("Tuition vs mobility") +
  xlab("Tuition costs ($)") +
  ylab("Mobility") +
  scale_x_continuous(labels = scales::dollar_format())

plot_two(p1, p2, title = "Tuition costs as a predictor of economic mobility")
```

```{r education corrplot}
cor_matrix <- cor(edu_mobility, use = "pairwise.complete.obs") 
corrplot(cor_matrix, 
         method = "color",
         type = "upper",
         col = colorRampPalette(c("tomato", "white", "dodgerblue"))(200),
         tl.cex = 0.6, 
         number.cex = 0.8,
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 30)
```

## Policy variables

```{r}
# Make a correlation matrix
cor_matrix <- cor(mobility, use = "pairwise.complete.obs")

# Make a heatmap
corrplot(cor_matrix,
         method = "color",  
         tl.col = "black",  
         tl.cex = 0.5, 
         tl.srt = 45,
         col = colorRampPalette(c("tomato", "white", "dodgerblue"))(200))       

# Pass the correlation matrix to a dataframe
cor_df <- as.data.frame(as.table(cor_matrix))
```

```{r}
# Remove diagonal correlations
cor_df <- cor_df %>%
  filter(Var1 != Var2)

# Standardize Var1 & Var2
cor_df <- cor_df %>%
  dplyr::rowwise() %>%
  dplyr::mutate(pair = paste(sort(c(Var1, Var2)), collapse = "_")) %>%
  dplyr::distinct(pair, .keep_all = TRUE) %>%
  dplyr::select(-pair)

# Sort by correlation
top_corr <- cor_df %>%
  arrange(desc(abs(Freq))) %>%  
  head(50)  

# Print strongest correlations
print(head(top_corr))
```
```{r}
# Create scatter plots
plot_scatter <- function(x_var, color, text_size = 5, r_p_size = 5, keep_axis_titles = FALSE) {  
  ggplot(mobility, aes(.data[[x_var]], School_spending)) +  
    geom_point(color = color, alpha = .3) + 
    geom_smooth(method = "lm", color = "black", se = FALSE) +  # Add linear regression line
    stat_cor(label.x = min(mobility[[x_var]], na.rm = TRUE), 
             label.y = max(mobility$School_spending, na.rm = TRUE) * 0.9, 
             size = r_p_size) +  # Adds R & p-values 
    ggtitle(paste(x_var)) + 
    theme_minimal() +
    theme(
      axis.title = if (keep_axis_titles) element_text(size = 10) else element_blank(),  
      axis.text = element_text(size = 8),  
      plot.title = element_text(hjust = 0.5, size = 10) 
    )
}

# Create plot vs mobility
mobility_plot <- plot_scatter("Mobility", "mediumseagreen", text_size = 5, r_p_size = 6, keep_axis_titles = TRUE) +
  ggtitle("Mobility vs School Spending")

# Create plots for other variables 
p1 <- plot_scatter("Seg_poverty", "mediumseagreen", r_p_size = 2)
p2 <- plot_scatter("Gini", "mediumseagreen", r_p_size = 2)
p3 <- plot_scatter("Gini_99", "mediumseagreen", r_p_size = 2)
p4 <- plot_scatter("Middle_class", "mediumseagreen", r_p_size = 2)
p5 <- plot_scatter("Single_mothers", "mediumseagreen", r_p_size = 2)
p6 <- plot_scatter("Test_scores", "mediumseagreen", r_p_size = 2)

# Display 
print(mobility_plot)

# Display all other plots on one page
grid.arrange(
  arrangeGrob(p1, p2, p3, p4, p5, p6, ncol = 2, 
              top = textGrob("Demographic Variables vs School Spending", 
                             gp = gpar(fontsize = 10, fontface = "bold")))
)
```
```{r, message=FALSE, warning=FALSE}
data <- mobility

# Create scatter plots
plot_scatter <- function(x_var, color, text_size = 5, r_p_size = 5, keep_axis_titles = FALSE) {  
  ggplot(data, aes(.data[[x_var]], Local_tax_rate)) + 
    geom_point(color = color, alpha = .3) + 
    geom_smooth(method = "lm", color = "black", se = FALSE) +  
    stat_cor(label.x = min(data[[x_var]]), 
             label.y = max(data$Local_tax_rate, na.rm = TRUE) * 0.9, 
             size = r_p_size) + 
    ggtitle(paste(x_var)) + 
    theme_minimal() +
    theme(
      axis.title = if (keep_axis_titles) element_text(size = 10) else element_blank(), 
      axis.text = element_text(size = 8),  
      plot.title = element_text(hjust = 0.5, size = 10) 
    )
}

# Create plot vs Mobility
mobility_plot <- plot_scatter("Mobility", "dodgerblue", text_size = 5, r_p_size = 6, keep_axis_titles = TRUE) +
  ggtitle("Mobility vs Local Tax Rate")

# Create plots for other variables 
p1 <- plot_scatter("Seg_poverty", "dodgerblue", r_p_size = 2)
p2 <- plot_scatter("Gini", "dodgerblue", r_p_size = 2)
p3 <- plot_scatter("Gini_99", "dodgerblue", r_p_size = 2)
p4 <- plot_scatter("Middle_class", "dodgerblue", r_p_size = 2)
p5 <- plot_scatter("Single_mothers", "dodgerblue", r_p_size = 2)
p6 <- plot_scatter("Test_scores", "dodgerblue", r_p_size = 2)

# Display 
print(mobility_plot)

# Arrange and display all other plots on one page 
grid.arrange(
  arrangeGrob(p1, p2, p3, p4, p5, p6, ncol = 2,  
              top = textGrob("Demographic Variables vs Local Tax Rate", 
                             gp = gpar(fontsize = 10, fontface = "bold")))
)
```

```{r, message=FALSE, warning=FALSE}
# Create scatter plots
plot_scatter <- function(x_var, color, text_size = 5, r_p_size = 5, keep_axis_titles = FALSE) {  
  ggplot(data, aes(.data[[x_var]], Local_gov_spending)) +  
    geom_point(color = color, alpha = .3, na.rm = TRUE) + 
    geom_smooth(method = "lm", color = "black", se = FALSE) +  
    stat_cor(label.x = min(data[[x_var]], na.rm = TRUE), 
             label.y = max(data$Local_gov_spending, na.rm = TRUE) * 0.9, 
             size = r_p_size) + 
    ggtitle(paste(x_var)) +  
    theme_minimal() +
    theme(
      axis.title = if (keep_axis_titles) element_text(size = 10) else element_blank(),  
      axis.text = element_text(size = 8),  
      plot.title = element_text(hjust = 0.5, size = 10)  
    )
}

# Create plot vs Mobility
mobility_plot <- plot_scatter("Mobility", "tomato", text_size = 5, r_p_size = 6, keep_axis_titles = TRUE) +
  ggtitle("Mobility vs Local Government Spending")

# Create plots for other variables 
p1 <- plot_scatter("Seg_poverty", "tomato", r_p_size = 2)
p2 <- plot_scatter("Gini", "tomato", r_p_size = 2)
p3 <- plot_scatter("Gini_99", "tomato", r_p_size = 2)
p4 <- plot_scatter("Middle_class", "tomato", r_p_size = 2)
p5 <- plot_scatter("Single_mothers", "tomato", r_p_size = 2)
p6 <- plot_scatter("Test_scores", "tomato", r_p_size = 2)

# Display 
print(mobility_plot)

# Arrange and display all other plots on one page 
grid.arrange(
  arrangeGrob(p1, p2, p3, p4, p5, p6, ncol = 2,  
              top = textGrob("Demographic Variables vs Local Government Spending", 
                             gp = gpar(fontsize = 10, fontface = "bold")))
)

```
```{r}
mobility[c("Local_tax_rate", "Local_gov_spending", "School_spending", "Mobility")] %>%
  ggpairs(aes(alpha = 0.5), 
          upper = list(continuous = wrap("cor", size = 3)),
          columnLabels = c("Tax Rate", "Gov Spending", "School Spending", "Mobility"),
          title = "Colinearity analysis of Government Policy",
          progress = FALSE)
```

## 
```{r}
mobility[c("Seg_poverty", "Seg_racial", "Seg_affluence", "Seg_income")] %>%
  ggpairs(aes(alpha = 0.5), 
          upper = list(continuous = wrap("cor", size = 5)),
          columnLabels = c("Poverty", "Race", "Affluence", "Income"),
          title = "Colinearity analysis of segregation",
          progress = FALSE)
```


While segregation on poverty lines is not particularly well correlated with segregation on racial lines, is is highly associated with segregation by affluence and segregation by income, which are also highly associated with each other. Since `Seg_poverty`, `Seg_affluence` and `Seg_income` are so strongly co-linear, `Seg_affluence` and `Seg_income` will be removed from the model. 

#### Income and income inequality

```{r}
mobility[c("Middle_class", "Income", "Gini", "Gini_99", "Share01")] %>%
  ggpairs(aes(alpha = 0.5), 
          upper = list(continuous = wrap("cor", size = 5)),
          columnLabels = c("Middle class", "Income", "Gini index", "Adj. Gini", "Share01"),
          title = "Colinearity analysis of income and income inequality",
          progress = FALSE)
```


The `Middle_class` variable is colinear with `Gini` and `Gini_99`, while the `Share01` variable is colinear with `Gini`. Additionally, `Gini` seems to be highly predictive of `Gini_99`. `Income` is not strongly associated with any of the other variables examined. 


### Social determinants of mobility

```{r warning=FALSE}
a <- ggplot(mobility, aes(x= Seg_racial)) + 
  geom_histogram(bins = 30, fill = 'dodgerblue', color = "gray30") + 
  ggtitle("Distribution of segregation by race")+
  xlab("Segregation by race") +
  ylab("Frequency")

b <- ggplot(data = mobility, aes(x = Mobility, y = Seg_racial)) + 
  geom_point(color = "dodgerblue", alpha = .3) + 
  stat_smooth(method = "lm", formula = y ~ x, geom = "line", color = "black", se = FALSE) +
  stat_cor(label.x=.17, label.y=.5) +
  ylim(0,.6) + 
  ggtitle("Segregation by race vs mobility") +
  xlab("Mobility") + 
  ylab("Racial segregation")

c <- ggplot(mobility, aes(x= Seg_poverty)) + 
  geom_histogram(bins = 30, fill = 'tomato', color = "gray30") + 
  ggtitle("Distribution of segregation by poverty")+
  xlab("Segregation by poverty") +
  ylab("Frequency") 

d <- ggplot(data = mobility, aes(x = Mobility, y = Seg_poverty)) + 
  geom_point(color = "tomato", alpha = .3) +
  stat_smooth(method = "lm", formula = y ~ x, geom = "line", color = "black", se = FALSE) +
  stat_cor(label.x=.17, label.y=.15) +
  ylim(0,.17) +
  ggtitle("Segregation by poverty v mobility") + 
  xlab("Mobility") + 
  ylab("Poverty segregation")

plot_four(a,c,b,d, title = "Segregation as a predictor of mobility")
```

