---
title: "RuthEDA"
author: "Ruth Walters"
date: "`r Sys.Date()`"
output: pdf_document
---
## Exploratory data analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Check which columns have NA/null values
print(colSums(is.na(mobility)))

```

```{r}
# Drop columns with >100 NA values
mobility <- mobility[,!(names(mobility) %in% c("Colleges","Tuition", "Graduation", "HS_dropout"))]

# Drop rows with NAs
mobility <- drop_na(mobility)
```

```{r}
# Recheck NA
for (i in colSums(is.na(mobility))) {
  if (as.numeric(i) != 0) {print(i)}
}
```

```{r}
mobility_numeric <- mobility[,!(names(mobility) %in% c("ID","Name", "State", "Latitude", "Longitude"))]

corrplot(cor(mobility_numeric),
         tl.col = "black",
         tl.cex = .5, 
         method = 'color')
```


Mobility appears to be highly positively correlated with the cluster of variables that measure segregation 

We can further identify three clusters of highly correlated variables:

- measures of segregation (`seg_racial`, `seg_income`, and `seg_affluence`)
- measures of the Gini index (`Gini`, `Share01`, `Gini_99` and `middle_class`)
- measures of migration (`migration_in` and `migration_out`)

```{r}
# Drop highly correlated variables
mobility <- mobility[,!(names(mobility) %in% c("Seg_income","Seg_affluence", "Share01", "Gini_99"))]

mobility_numeric <- mobility[,!(names(mobility) %in% c("ID","Name", "State", "Latitude", "Longitude"))]
```

```{r}
ggplot(mobility, aes(Mobility, Gini)) + 
  geom_point() +
  facet_wrap(~State)
```