---
title: "RuthEDA"
author: "Ruth Walters"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r}
# Import dependencies
library(ggplot2)
library(dplyr)
library(corrplot)
library(tidyr)

theme_set(theme_bw())

# Import data
mobility <- read.csv("mobility-all.csv", header = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
mobility <- read.csv("mobility-all.csv", header = TRUE)

library(ggplot2)
library(dplyr)
library(corrplot)
library(tidyr)
library(cowplot)
library(ggpubr)
library(GGally)

theme_set(theme_bw())
theme_update(
  axis.title.x = element_blank(),  
  axis.title.y = element_blank(), 
  plot.title = element_text(size = 12, face = "italic")
)
```
## From datacamp
[https://www.datacamp.com/tutorial/linear-regression-R][How to Do Linear Regression in R]
 
> "When a regression takes into account two or more predictors to create the linear regression, it’s called multiple linear regression. In R, to add another coefficient, add the symbol "+" for every additional variable you want to add to the model.

Linear model: `lm([target] ~ [predictor], data = [data source])`

## Data preparation
```{r}
mobility <- read.csv("mobility-all.csv", header = TRUE, stringsAsFactors = TRUE)
```

Drop all non-quantitative rows
```{r}
quals <- c("ID","Name", "State", "Latitude", "Longitude")

mobility <- mobility[,!(names(mobility) %in% quals)]
```

Drop low-quality columns
```{r}
print(colSums(is.na(mobility)))
```

```{r}
bad_cols <- c("Colleges","Tuition", "Graduation", "HS_dropout") # +100 NULL
mobility <- mobility[,!(names(mobility) %in% bad_cols)]
```

Drop remaining NULLS
```{r}
before <- nrow(mobility)
mobility <- drop_na(mobility)
dropped <- before - nrow(mobility)

print("Data reduced by: ")
print((dropped/before))
```

## Exploratory data analysis
```{r}
corrplot(cor(mobility),
         tl.col = "black",
         tl.cex = .5, 
         method = 'color')
```

### Explore highly correlated variables
#### Segregation


```{r}
mobility[c("Seg_poverty", "Seg_racial", "Seg_affluence", "Seg_income")] %>%
  ggpairs(aes(alpha = 0.5), 
          upper = list(continuous = wrap("cor", size = 5)),
          columnLabels = c("Poverty", "Race", "Affluence", "Income"),
          title = "Colinearity analysis of segregation",
          progress = FALSE)
```


While segregation on poverty lines is not particularly well correlated with segregation on racial lines, is is highly associated with segregation by affluence and segregation by income, which are also highly associated with each other. Since `Seg_poverty`, `Seg_affluence` and `Seg_income` are so strongly co-linear, `Seg_affluence` and `Seg_income` will be removed from the model. 

#### Income and income inequality

```{r}
mobility[c("Middle_class", "Income", "Gini", "Gini_99", "Share01")] %>%
  ggpairs(aes(alpha = 0.5), 
          upper = list(continuous = wrap("cor", size = 5)),
          columnLabels = c("Middle class", "Income", "Gini index", "Adj. Gini", "Share01"),
          title = "Colinearity analysis of income and income inequality",
          progress = FALSE)
```


The `Middle_class` variable is colinear with `Gini` and `Gini_99`, while the `Share01` variable is colinear with `Gini`. Additionally, `Gini` seems to be highly predictive of `Gini_99`. `Income` is not strongly associated with any of the other variables examined. 

### Explore non-linear variables
```{r}
mobility %>%
  gather(-Mobility, key = "var", value = "value") %>%
  ggplot(aes(x = value, y = Mobility)) +
  geom_point(alpha = 0.3) +
  facet_wrap(~ var, scales = "free")

```

### Social determinants of mobility

```{r}
a <- ggplot(data = mobility, aes(x = Mobility, y = Seg_racial)) + 
  geom_point(color = "cornflowerblue", alpha = .3) + 
  #stat_smooth(method = "lm", formula = y ~ x, geom = "line", color = "darkorange") +
  stat_cor(label.x=.17, label.y=.5) +
  ggtitle("Race")
    
# b <- ggplot(data = mobility, aes(x = Mobility, y = Seg_income)) + 
#   geom_point(color = "skyblue", alpha = .3) +
#   stat_cor(label.x=.17, label.y=.12) +
#   ggtitle("Income")

c <- ggplot(data = mobility, aes(x = Mobility, y = Seg_poverty)) + 
  geom_point(color = "mediumseagreen", alpha = .3) +
  stat_cor(label.x=.17, label.y=.15) +
  ylim(0,.17) +
  ggtitle("Poverty") + 
  xlab("Mobility") + 
  ylab("Segregation") +
  theme(axis.title.x = element_text(hjust = 0),
        axis.title.y = element_text(angle=90, hjust = 0, margin = margin(r = 5)))

# d <- ggplot(data = mobility, aes(x = Mobility, y = Seg_affluence)) + 
#   geom_point(color = "seagreen", alpha = .3) +
#   stat_cor(label.x=.17, label.y=.15) +
#   ylim(0,.17) +
#   ggtitle("Affluence") 

plot_row <- plot_grid(c,a, align = "hv")

title <- ggdraw() + 
  draw_label(
    "Segregation as a predictor of mobility",
    fontface = 'bold',
    x = 0,
    hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 7))
plot_grid(
  title, plot_row,
  ncol = 1,
  rel_heights = c(0.1, 1)
)
```

```{r warning=FALSE}
mobility$Urban <- as.factor(mobility$Urban)

theme_update(
  axis.title.x = element_text(hjust = 0),  
  axis.title.y = element_text(element_text(angle=90, hjust = 0, margin = margin(r = 5)))) 

a <- ggplot(data = mobility, aes(x = Progressivity, y = Mobility)) + 
  geom_smooth() + 
  ggtitle("Progressivity") + 
  xlab("Progressivity") +
  ylab(" ")

b <- ggplot(data = mobility, aes(x = Urban, y = Mobility)) + 
  geom_bar(position='dodge', stat='summary', fun='mean', alpha = .3) +
  ggtitle("Urban communities") + 
  xlab("Community type") + 
  ylab("Average mobility")

c <- ggplot(data = mobility, aes(x = Single_mothers, y = Mobility)) + 
  geom_point(alpha = .3, pch = 16) +
  stat_cor(label.x=.17, label.y=.3, label.size = 0.01) +
  ggtitle("Proportion of single mothers") +
  xlab("p Single mothers") +
  ylab("Mobility")

d <- ggplot(data = mobility, aes(x = Violent_crime, y = Mobility)) + 
  geom_violin() +
  stat_cor(label.x=.0031, label.y=.3, label.size = 0.001) +
  xlab("Violent crime") + 
  ylab("Mobility") + 
  ggtitle("Violent crime incidence")

plot_row <- plot_grid(a,b,c,d, align = "none")

title <- ggdraw() + 
  draw_label(
    "Community factors associated with mobility",
    fontface = 'bold',
    x = 0,
    hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 7))
plot_grid(
  title, plot_row,
  ncol = 1,
  rel_heights = c(0.1, 1)
)
```
